Context-dependent consequences of lagged effects in demographic models

### 2025-09-08 docs/ipm_comparison_ms.qmd
updates - eric comments

---
title: "Context-dependent consequences of lagged effects in demographic models"
shorttitle: "Lagged Effects IPMs"  
authors:
  - name: Eric R. Scott
    email: scottericr@gmail.com
    affiliations:
      - ref: 1
    corresponding: true
  - name: María Uriarte
    email: mu2126@columbia.edu
    affiliations:
      - ref: 2
  - name: Emilio M. Bruna
    email: embruna@ufl.edu
    affiliations:
      - ref: 1
      - ref: 3
      - ref: 4
affiliations:
  - id: "1"
    name: Department of Wildlife Ecology and Conservation, University of Florida, Gainesville, Florida 32611-0430 USA
  - id: "2" 
    name: Department of Ecology, Evolution and Environmental Biology, Columbia University 1200 Amsterdam Avenue, New York, New York 10027 USA
  - id: "3"
    name: Center for Latin American Studies, University of Florida, Gainesville, Florida 32611-5530 USA
  - id: "4"
    name: Biological Dynamics of Forest Fragments Project, INPA-PDBFF, CP 478, Manaus, Amazonas 69011-970 Brazil
date: 'last-modified'
date-format: '([draft:] D MMMM YYYY)'
linkcolor: blue
bibliography: lagged-ipms-ms.bib
csl: "templates/the-american-naturalist.csl" # Insert path for the bib-style
# 
# highlights: |
#   These are the highlights
format: 
  pdf:
    # template: AmNat_MS_template.tex
    # documentclass: scrarticle
    documentclass: article
    number-sections: false
    colorlinks: true
    keeptex: true
    fontsize: 12pt
    # the following line is to render the aaffiliations in pdf, which is not default in quarto, see 
    # https://stackoverflow.com/questions/75040607/why-do-affiliations-not-show-up-anywhere-in-the-pdf-output-of-quarto
    template-partials:
      - title.tex
    include-in-header:
      text: |
        \usepackage[noblocks]{authblk}
        \renewcommand*{\Authsep}{, }
        \renewcommand*{\Authand}{, }
        \renewcommand*{\Authands}{, }
        \renewcommand\Affilfont{\normalsize}
        \usepackage{lineno}\linenumbers
        \usepackage{sectsty}\sectionfont{\centering}
        \usepackage{sectsty}\subsectionfont{\centering}
        \usepackage{fontspec}
        \usepackage{titlesec}
        <!-- \usepackage[sc]{mathpazo} %Like Palatino with extensive math support -->
        <!-- \usepackage[utf8]{inputenc} -->
        \usepackage{parskip}
        \setlength{\parskip}{1pt} % 1ex plus 0.5ex minus 0.2ex}
        \setlength{\parindent}{0pt}
        \usepackage{caption}
        \captionsetup[figure]{labelfont={bf},name={fig.},labelsep=period}
        \raggedright
        \usepackage{setspace}\doublespacing
        \setlength\parindent{22pt}

@@

Current environmental conditions can have large and immediate effects on the growth, survival, or reproduction of long-lived organisms. There is also mounting evidence, however, for intervals that range from months to years between environmental conditions and the resulting changes in demographic vital rates. These *Lagged Effects*, also known as *Delayed Life-history Events* (i.e., DLHEs) [@beckermanPopulationDynamicConsequences2002], can simultaneously affect an entire cohort (e.g., juveniles hatching during a period of scarcity will all have delayed maturation and lower lifetime fecundity) or only a subset of the population (e.g., cold temperatures in one year lead to reduced flowering by potentially reproductive individuals in the next). In addition, the temporal delay between an environmental event and changes in demographic vital rates depends on both the intensity of the event and its timing relative to the underlying physiological processes [@eversLaggedDormantSeason2021; @crileyYearProductionHigh1994]. A drought during the early stages of gestation or floral bud formation, for example, might have a much larger impact on the number of fruits or offspring produced than one at a later stage of the reproductive process. The delay or intensity of lagged effects can also be location- or habitat-specific, with individuals in some sites or habitats buffered against -- or able to recover more quickly from - the delayed effects of environmental variation.

Because Lagged Effects are often directly linked to reproduction and survival, it has been argued they could have major but underappreciated consequences for population dynamics [@beckermanPopulationDynamicConsequences2002]. While recent studies suggest this could indeed the case [e.g., @molowny-horasChangesNaturalDynamics2017; @tenhumbergTimelaggedEffectsWeather2018; @williamsLifeHistoryEvolution2015], broader efforts to test this hypothesis have face a number of obstacles [@metcalfStatisticalModellingAnnual2015]. Empirical tests of putative lagged effects because these are challenging to design, implement, and maintain [@kussEvolutionaryDemographyLonglived2008a]. While one could use conceivably use observational data to detect lagged effects, doing so requires long-term data on both the putative lagged effect (e.g., reduced probability of flowering) and its potential environmental drivers (e.g., drought during early floral development), and such coupled data sets are rare [sensu @eversLaggedDormantSeason2021]. In addition, the methods for both identifying lagged effects and modeling their demographic impacts can be challenging to implement. For example, many of the potentially applicable statistical methods have stringent assumptions and data requirements rarely met by ecological data [@metcalfStatisticalModellingAnnual2015], while the including complex biological processes in demographic models can render them less tractable or amenable to analysis using currently available tools. Addressing these challenges is a major undertaking; the value of doing so will depend on the effort required vs. the potential consequences of failing to consider lagged effects - consequences that range from overestimating projections of population growth rate (i.e., $\lambda$) in a conservation setting to drawing invalid conclusions regarding support for the predictions of ecological or evolutionary theory.

Integral Projection Models (i.e., IPMs) are an important and widely used tool for studying demography and population dynamics [@ellnerIntegralProjectionModels2006; @reesBuildingIntegralProjection2014; @reesIntegralProjectionModels2009]. Their flexibility, in concert with a rapidly growing suite of software, data, and other resources [@ellnerDatadrivenModellingStructured2016; @levinIpmrFlexibleImplementation2021; @salguero-gomezCOMPADREPlantMatrix2015], have facilitated their use to study a wide range of topics in ecology, evolution, and conservation [@ellnerDatadrivenModellingStructured2016; @morrisQuantitativeConservationBiology2002 and @croneHowPlantEcologists2011]. Mathematical and statistical advances [e.g., @williams2012avoiding; @brooksStatisticalModelingPatterns2019] have rapidly expand the scope of questions and biological processes that can be investigated with these models [e.g., @metcalfStatisticalModellingAnnual2015; @reesEvolvingIntegralProjection2016; @ellnerDatadrivenModellingStructured2016]. Here we investigate how including lagged effects in Integral Projection Models influences projections of $\lambda$ and population structure.

We have previously shown that the effects of precipitation extremes on the demographic vital rates of an Amazonian understory herb (*Heliconia acuminata*, Heliconiaceae) can be delayed up to 36 months [@scottDelayedEffectsClimate2022], with the presence and duration of these lagged effects varying by vital rate and habitat. We parameterized three classes of Integral Projection Models - a deterministic IPM, a stochastic IPM, and a stochastic IPM with lagged effects of precipitation on vital rates - for populations in two habitat type (i.e., continuous forest vs. forest fragments). Based on previous studies [@brunaArePlantPopulations2003; @brunaDemographicEffectsHabitat2005; @brunaHabitatFragmentationDemographic2002] and demographic theory [@caswell2001; @tuljapurkarPopulationDynamicsVariable1990] we predicted that: (i) projections of $\lambda$ from deterministic models would be higher than those of stochastic models, and that (ii) $\lambda$ is lower for Forest Fragment than Continuous Forest populations regardless of model, but the difference between the two habitats is greatest for the IPM with lagged effects.

@@

*Heliconia acuminata* (Heliconiaceae) is a perennial, self-incompatible monocot that is distributed throughout much of the Amazon basin [@kressDiversityDistributionHeliconia1990]. While some *Heliconia* species grow in large aggregations on roadsides, gaps, and in other disturbed habitats, others - including *H. acuminata* - grow primarily in the forest understory [@kressSelfincompatibilitySystemsCentral1983; @ribeiroInfluencePostclearingTreatment2010]. Understory *Heliconia* species produce fewer flowers and are pollinated by traplining hummingbirds [@brunaHeliconiaAcuminataReproductive2004; @stoufferForestFragmentationSeasonal1996]. The models and analyses here are based on 11 years (1998-2009) of demographic data collected on \>8500 *H. acuminata* found at Brazil's Biological Dynamics of Forest Fragments Project (BDFFP), located \~70 km north of Manaus, Brazil. The BDFFP reserves include both continuous forest and forest fragments that range in size from 1-100 ha. These fragment reserves were originally isolated in the early 1980’s by the creation of cattle pastures, with the secondary growth surrounding them periodically cleared to ensure their continued isolation. The habitat in all sites is non-flooded lowland rain forest with rugged topography. A complete summary of the BDFFP and its history can be found in @bierregaardLessonsAmazoniaEcology2001.\

@@

*Heliconia acuminata* in our site begin flowering early in the rainy season (e.g., January) and most reproductive plants produce a single infloresence (range = 1--7) with 20--25 flowers [@brunaHabitatFragmentationDemographic2002]. Fruits mature April-May and have 1--3 seeds per fruit ($\bar{x}=2$) that are dispersed by a thrush and several species of manakin [@uriarteDisentanglingDriversReduced2011]. Dispersed seeds germinate approximately 6 months after dispersal at the onset of the subsequent rainy season, with rates of germination and seedling establishment higher in continuous forest than forest fragments [@brunaEffectsForestFragmentation2002; @brunaSeedGerminationRainforest1999]. On average plots in CF also had more than twice as many plants as the plots in 1-ha fragments (CF median = 788, range = (201-1549); 1-ha median = 339, range = (297-400)).

@@

We projected the growth rate and structure of *Heliconia acuminata* with three classes of IPMs - Deterministic, Stochastic, and Stochastic with Lagged Environmental Effects. Each of these IPMs required different functional forms of the underlying vital rate functions used to describe the *H. acuminata* life cycle (@fig-lifecycle). All models were density-independent, with the deterministic model serving as the foundation for the more complex models. Our modeling workflows were managed using the `targets` R package [@targets]; in adddition to ensuring reproducibility this allowed us to track computational time spent processing and analyzing each class of IPM.\

@@

We used the annual census data [@brunaArePlantPopulations2003] to fit the deterministic vital rate functions for growth, survival, and flowering in each habitat type (i.e., Fragments, Continuous Forest; the data from all plots within a habitat class were pooled to create a single 'summary population' for the CF and FF habitats [@brunaArePlantPopulations2003]. For established plants these were modeled as a smooth function of size in the previous census with generalized additive models (GAMs) fit with the `mgcv` library [@mgcv] for the R statistical programming language [@rcoreteam2020].<!--`r version$version.string`---> <!--# probably include a supplement with detail on things like basis, number of knots, etc. --> For consistency, seedling survival and growth were also modeled using GAMs, but without size in the previous census as a predictor (i.e. 'intercept-only' models). For growth models a scaled t family distribution provided a better fit to the data than a Gaussian fit, as the residuals with a simple Gaussian model were Leptokurtic. To model size-specific fecundity we used the data on fruits per flowering plant [@brunaLeafNumberLeaf2021], seeds per fruit [@brunaDatasetHeliconiaAcuminata2014], and the experimentally-derived estimates of seed germination and seedling establishment [@brunaEffectsForestFragmentation2002; @brunaSeedGerminationRainforest1999]. <!--# Emilio: add more info here -->\

@@

**(3) Stochastic IPMs with lagged effects of precipitation on vital rates:** We explicit modeled the lagged effects of precipitation extremes on vital rates using the procedure described in @scottDelayedEffectsClimate2022. Briefly, we first calculated the Standardized Precipitation Evapotranspiraton Index (i.e., SPEI) for our study site using a published gridded dataset based on ground measurements [@xavierDailyGriddedMeteorological2016]. After we fit vital rate models using the long-term survey data (@fig-lifecycle), we modeled delayed effects of SPEI with Distributed Lag Non-linear Models (i.e., DLNMs) with a maximum lag of 36 months [@scottDelayedEffectsClimate2022]. To iterate these parameter-resampled IPMs [*sensu* @metcalfStatisticalModellingAnnual2015] we first created a random sequence of SPEI values by sampling years of the observed (monthly) SPEI data. For every year we then calculated a lag of 36 months from the month in which that year's census was completed. These values were then used to predict the fitted values from vital rate models, which generated different kernels for each iteration of the IPM. Though the kernels of successive iterations are not entirely independent -- the SPEI values used to calculate vital rates include values used in the previous two iterations -- they are ergodic.\

@@

We used a Generalized Linear Model to compare the projections of $\lambda$ from the two stochastic IPMs. We modelled the lambda as a function of IPM type (kernel-resampling vs parameter-resampling) and Habitat (continuous forest vs. fragments), and the interaction of IPM type and Habitat. Because the response variable ($\lambda$) was continuous we used the Gaussian distribution with the identity function in our model. We verified the model assumptions by plotting residuals versus fitted values.\

@@

Finally, we compared the structure of populations through the first 250 time steps in each habitat by each of the stochastic IPMs (i.e, kernel- vs. parameter-resampling). At each time step, we assigned the individuals to one of four stage classes based on plant size, probability of survival, and probability of reproduction (see Table @tbl-plantcats). We then calculated the proportion of the population that was in each stage class in each habitat for each IPM type. Comparing the means and variances of these proportions will allow us to determine (1) if the two IPMs project different population structures over time, (2) if the population structure for a given IPM is consistent across habitat type. Because Shapiro-Wilk tests indicated that the distributions of proportions were not normally distributed, we used the non-parametric Ansari-Bradley Test to compare the distribution of proportions in each habitat x stage combination. These analyses were conducted using the R libraries `rstatix` and `vartest`, respectively [@kassambaraRstatixPipefriendlyFramework2023; @cosarVartestTestsVariance2024].

@@

All IPMs projected higher population growth rates for *Heliconia acuminata* populations in Continuous Forest than for those in Forest Fragments (@tbl-lambdas, @tbl-glm). The differences between $\lambda^{CF}$ and $\lambda^{FF}$, which ranged from 1.19-2%, were significant for both deterministic ($P[\theta \geq \theta_{obs}]$ = `r det_stats$p_val`, N = 1000 randomization) and stochastic IPMs (@tbl-glm). There were also significant effects of IPM Type, Habitat, and their interaction on $\lambda$ (@tbl-glm). Deterministic $\lambda$ was nearly identical to the average $\lambda$ from kernel-resampled IPMs (@fig-lambdas), which is somewhat surpring given that stochasticity is predicted to reduce population growth rates [@doakCorrectlyEstimatingHow2005a; @tuljapurkarManyGrowthRates2003; @metcalfStatisticalModellingAnnual2015]. However including lagged effects in IPMs resulted in projections of $\lambda$ that were on average 1.5-2% lower than those of deterministic models, with 70-76% of projections below deterministic $\lambda$ (@fig-lambdas). The IPMs with lagged effects also appear to be more accurate - the underlying vital rate models for IPMs with lagged effects all had the best fit to the survey data (dAIC = 0, @tbl-aic). <!--# No longer true if random effect = different smooth for each year -->

@@

What are the implications of these results for the use of IPMs to study plant demography? First, they suggest that failing to include lagged effects in models could result in underestimates of $\lambda$. While relative differences in $\lambda$ may be sufficient for some tests of theory, they may be critical in conservation or management settings because even small differences in $\lambda$ can have major consequences for population dynamics. They could also be critical for studies assessing the effects of climate change, given many DLHE are climate driven. That said, there is a cost to including lagged effects in models, even in if they are more accurate. While Deterministic and Kernel-resampled Stochastic IPMs took only \~`r det_time` and \~`r sker_time` min to iterate (respectively), the Stochastic IPMs with using parameter-resampled kernels and lagged effects took \~`r spar_time` min. This is largely due to the required computational resources and algorithms (i.e., `predict()`), which are much slower forGeneral Additive Models (i.e., GAMs) with 2D smooths because of the much higher number of knots than for the GAMs used in Determinsitic and Kernel-resampled IPMs. Advances in computational power and access to high-performance computing resources could lower this cost.

Computational power cannot compensate for limited data, however. Detecting lagged effects and evaluating their consequences requires long-term demographic data - data that are only available for a relatively small number of species, few of which are in the tropics. The increasing evidence that lagged effects are ubiquitous, and that they can have major demographic impacts, underscores the need to support the collection of such long-term data, the complementary development of experimental and statistical approaches to disentangling lagged effects, and community driven efforts to identify priority or model systems for in-depth investigation.\

@@

@@

<!-- \newpage -->

<!-- ```{r #fig-var,  echo=FALSE, out.width="85%",fig.align='center'} -->

<!-- #| fig.cap = "------" -->

<!-- # The figures is currently generated with "plotting_lambdas.R, need to move to targets_r  -->

<!-- path <- here("docs","figures","relative_variances_fig.png") -->

<!-- knitr::include_graphics(path, rel_path = TRUE) -->

<!-- \newpage -->

<!-- ```{r #fig-marginal-long,  echo=FALSE, out.width="85%",fig.align='center'} -->

<!-- #| fig.cap = "The relative proportion of the population in each size class (FF vs. CF) for each of 250 iterations of the kernel-resampled (dark shading) and parameter-resampled IPM models (light shading). Values on the 1-1 line indicate an iteration where CF and FF have the same proportion of the population in a given size class; the marginal plots indicate the distribution of these relative proportions for each class of IPM in each habitat. Note that both the scatterplots and marginal plots exclude transient dynamics (iterations 1-30)." -->

<!-- # The figures is currently generated with "plotting_lambdas.R, need to move to targets_r  -->

<!-- path <- here("docs","figures","marginal_stage_fig_long.png") -->

<!-- knitr::include_graphics(path, rel_path = TRUE) -->

>>>>

---
title: "Context-dependent consequences of lagged effects in demographic models"
shorttitle: "Lagged Effects IPMs"  
authors:
  - name: Eric R. Scott
    email: scottericr@gmail.com
    affiliations:
      - ref: 1
    corresponding: true
  - name: María Uriarte
    email: mu2126@columbia.edu
    affiliations:
      - ref: 2
  - name: Emilio M. Bruna
    email: embruna@ufl.edu
    affiliations:
      - ref: 1
      - ref: 3
      - ref: 4
affiliations:
  - id: "1"
    name: Department of Wildlife Ecology and Conservation, University of Florida, Gainesville, Florida 32611-0430 USA
  - id: "2" 
    name: Department of Ecology, Evolution and Environmental Biology, Columbia University 1200 Amsterdam Avenue, New York, New York 10027 USA
  - id: "3"
    name: Center for Latin American Studies, University of Florida, Gainesville, Florida 32611-5530 USA
  - id: "4"
    name: Biological Dynamics of Forest Fragments Project, INPA-PDBFF, CP 478, Manaus, Amazonas 69011-970 Brazil
date: 'last-modified'
date-format: '([draft:] D MMMM YYYY)'
linkcolor: blue
bibliography: lagged-ipms-ms.bib
csl: "templates/the-american-naturalist.csl" # Insert path for the bib-style
# 
# highlights: |
#   These are the highlights
format: 
  pdf:
    # template: AmNat_MS_template.tex
    # documentclass: scrarticle
    documentclass: article
    number-sections: false
    colorlinks: true
    keeptex: true
    fontsize: 12pt
    fig-pos: "H"
    # the following line is to render the aaffiliations in pdf, which is not default in quarto, see 
    # https://stackoverflow.com/questions/75040607/why-do-affiliations-not-show-up-anywhere-in-the-pdf-output-of-quarto
    template-partials:
      - title.tex
    include-in-header:
      text: |
        \usepackage[noblocks]{authblk}
        \renewcommand*{\Authsep}{, }
        \renewcommand*{\Authand}{, }
        \renewcommand*{\Authands}{, }
        \renewcommand\Affilfont{\normalsize}
        \usepackage{lineno}\linenumbers
        \usepackage{sectsty}\sectionfont{\centering}
        \usepackage{sectsty}\subsectionfont{\centering}
        \usepackage{fontspec}
        \usepackage{titlesec}
        <!-- \usepackage[sc]{mathpazo} %Like Palatino with extensive math support -->
        <!-- \usepackage[utf8]{inputenc} -->
        \usepackage{parskip}
        \setlength{\parskip}{1pt} % 1ex plus 0.5ex minus 0.2ex}
        \setlength{\parindent}{0pt}
        \usepackage{caption}
        \captionsetup[figure]{labelfont={bf},name={fig.},labelsep=period}
        \raggedright
        \usepackage{setspace}\doublespacing
        \setlength\parindent{22pt}

@@

Current environmental conditions can have large and immediate effects on the growth, survival, or reproduction of long-lived organisms. There is also mounting evidence, however, for intervals that range from months to years between environmental conditions and the resulting changes in demographic vital rates. These *Lagged Effects*, also known as *Delayed Life-history Events* (i.e., DLHEs) [@beckermanPopulationDynamicConsequences2002], can simultaneously affect an entire cohort (e.g., juveniles hatching during a period of scarcity will all have delayed maturation and lower lifetime fecundity) or only a subset of the population (e.g., cold temperatures in one year lead to reduced flowering by potentially reproductive individuals in the next). In addition, the temporal delay between an environmental event and changes in demographic vital rates depends on both the intensity of the event and its timing relative to the underlying physiological processes [@eversLaggedDormantSeason2021; @crileyYearProductionHigh1994]. A drought during the early stages of gestation or floral bud formation, for example, might have a much larger impact on the number of fruits or offspring produced than one at a later stage of the reproductive process. The delay or intensity of lagged effects can also be location- or habitat-specific, with individuals in some sites or habitats buffered against, or able to recover more quickly from, the delayed effects of environmental variation.

Because Lagged Effects are often directly linked to reproduction and survival, it has been argued they could have major but underappreciated consequences for population dynamics [@beckermanPopulationDynamicConsequences2002]. While recent studies suggest this could indeed the case [e.g., @molowny-horasChangesNaturalDynamics2017; @tenhumbergTimelaggedEffectsWeather2018; @williamsLifeHistoryEvolution2015], broader efforts to test this hypothesis have face a number of obstacles [@metcalfStatisticalModellingAnnual2015]. Empirical tests of putative lagged effects are rare because these are challenging to design, implement, and maintain [@kussEvolutionaryDemographyLonglived2008a]. While one could conceivably use observational data to detect lagged effects, doing so requires long-term data on both the putative lagged effect (e.g., reduced probability of flowering) and its potential environmental drivers (e.g., drought during early floral development), and such coupled data sets are rare [sensu @eversLaggedDormantSeason2021]. In addition, the methods for both identifying lagged effects and modeling their demographic impacts can be challenging to implement. For example, many of the potentially applicable statistical methods have stringent assumptions and data requirements rarely met by ecological data [@metcalfStatisticalModellingAnnual2015], while the including complex biological processes in demographic models can render them less tractable or amenable to analysis using currently available tools. Addressing these challenges is a major undertaking; the value of doing so will depend on the effort required vs. the potential consequences of failing to consider lagged effects --- consequences that range from overestimating projections of population growth rate (i.e., $\lambda$) in a conservation setting to drawing invalid conclusions regarding support for the predictions of ecological or evolutionary theory.

Integral Projection Models (i.e., IPMs) are an important and widely used tool for studying demography and population dynamics [@ellnerIntegralProjectionModels2006; @reesBuildingIntegralProjection2014; @reesIntegralProjectionModels2009]. Their flexibility, in concert with a rapidly growing suite of software, data, and other resources [@ellnerDatadrivenModellingStructured2016; @levinIpmrFlexibleImplementation2021; @salguero-gomezCOMPADREPlantMatrix2015], have facilitated their use to study a wide range of topics in ecology, evolution, and conservation [@ellnerDatadrivenModellingStructured2016; @morrisQuantitativeConservationBiology2002; @croneHowPlantEcologists2011]. Mathematical and statistical advances [e.g., @williams2012avoiding; @brooksStatisticalModelingPatterns2019] have rapidly expanded the scope of questions and biological processes that can be investigated with these models [e.g., @metcalfStatisticalModellingAnnual2015; @reesEvolvingIntegralProjection2016; @ellnerDatadrivenModellingStructured2016]. Here we investigate how including lagged effects in Integral Projection Models influences projections of $\lambda$ and population structure.

We have previously shown that the effects of precipitation extremes on the demographic vital rates of an Amazonian understory herb (*Heliconia acuminata*, Heliconiaceae) can be delayed up to 36 months [@scottDelayedEffectsClimate2022], with the presence and duration of these lagged effects varying by vital rate and habitat. We parameterized three classes of Integral Projection Models --- a deterministic IPM, a stochastic IPM, and a stochastic IPM with lagged effects of precipitation on vital rates --- for populations in two habitat types (i.e., continuous forest vs. forest fragments). Based on previous studies [@brunaArePlantPopulations2003; @brunaDemographicEffectsHabitat2005; @brunaHabitatFragmentationDemographic2002] and demographic theory [@caswell2001; @tuljapurkarPopulationDynamicsVariable1990] we predicted that: (i) projections of $\lambda$ from deterministic models would be higher than those of stochastic models, and that (ii) $\lambda$ is lower for Forest Fragment than Continuous Forest populations regardless of model, but the difference between the two habitats is greatest for the IPM with lagged effects.

@@

*Heliconia acuminata* (Heliconiaceae) is a perennial, self-incompatible monocot that is distributed throughout much of the Amazon basin [@kressDiversityDistributionHeliconia1990]. While some *Heliconia* species grow in large aggregations on roadsides, gaps, and in other disturbed habitats, others, including *H. acuminata*, grow primarily in the forest understory [@kressSelfincompatibilitySystemsCentral1983; @ribeiroInfluencePostclearingTreatment2010]. Understory *Heliconia* species produce fewer flowers and are pollinated by traplining hummingbirds [@brunaHeliconiaAcuminataReproductive2004; @stoufferForestFragmentationSeasonal1996]. The models and analyses here are based on 11 years (1998-2009) of demographic data collected on \>8500 *H. acuminata* found at Brazil's Biological Dynamics of Forest Fragments Project (BDFFP), located \~70 km north of Manaus, Brazil. The BDFFP reserves include both continuous forest and forest fragments that range in size from 1-100 ha. These fragment reserves were originally isolated in the early 1980’s by the creation of cattle pastures, with the secondary growth surrounding them periodically cleared to ensure their continued isolation. The habitat in all sites is non-flooded lowland rain forest with rugged topography. A complete summary of the BDFFP and its history can be found in @bierregaardLessonsAmazoniaEcology2001.\

@@

The rainy season in our sites is typically from late December through late May. *Heliconia acuminata* in our site begin flowering early in the rainy season and most reproductive plants produce a single infloresence (range = 1--7) with 20--25 flowers [@brunaHabitatFragmentationDemographic2002]. Fruits mature April-May and have 1--3 seeds per fruit ($\bar{x}=2$) that are dispersed by a thrush and several species of manakin [@uriarteDisentanglingDriversReduced2011]. Dispersed seeds germinate approximately 6 months after dispersal at the onset of the subsequent rainy season, with rates of germination and seedling establishment higher in continuous forest than forest fragments [@brunaEffectsForestFragmentation2002; @brunaSeedGerminationRainforest1999]. On average, plots in CF also had more than twice as many plants as the plots in 1-ha fragments (CF median = 788, range = (201-1549); 1-ha median = 339, range = (297-400)).

@@

We projected the growth rate and structure of *Heliconia acuminata* with three classes of IPMs: Deterministic, Stochastic, and Stochastic with Lagged Environmental Effects. Each of these IPMs required different functional forms of the underlying vital rate functions used to describe the *H. acuminata* life cycle (@fig-lifecycle). All models were density-independent, with the deterministic model serving as the foundation for the more complex models. Our modeling workflows were managed using the `targets` R package [@targets]; in adddition to ensuring reproducibility this allowed us to track computational time spent processing and analyzing each class of IPM.\

@@

We used the annual census data [@brunaArePlantPopulations2003] to fit the deterministic vital rate functions for growth, survival, and flowering in each habitat type (i.e., Fragments, Continuous Forest; the data from all plots within a habitat class were pooled to create a single 'summary population' for the CF and FF habitats [@brunaArePlantPopulations2003]. For established plants these were modeled as a smooth function of size in the previous census with generalized additive models (GAMs) fit with the `mgcv` package [@mgcv] for the R statistical programming language [@rcoreteam2020].<!--`r version$version.string`---> <!--# probably include a supplement with detail on things like basis, number of knots, etc. --> For consistency, seedling survival and growth were also modeled using GAMs, but without size in the previous census as a predictor (i.e. 'intercept-only' models). For growth models a scaled t family distribution provided a better fit to the data than a Gaussian fit, as the residuals with a simple Gaussian model were Leptokurtic. To model size-specific fecundity we used the data on fruits per flowering plant [@brunaLeafNumberLeaf2021], seeds per fruit [@brunaDatasetHeliconiaAcuminata2014], and the experimentally-derived estimates of seed germination and seedling establishment [@brunaEffectsForestFragmentation2002; @brunaSeedGerminationRainforest1999]. <!--# Emilio: add more info here -->\

@@

**(3) Stochastic IPMs with lagged effects of precipitation on vital rates:** We explicit modeled the lagged effects of precipitation extremes on vital rates using the procedure described in @scottDelayedEffectsClimate2022. Briefly, we first calculated the Standardized Precipitation Evapotranspiraton Index (i.e., SPEI) for our study site using a published gridded dataset based on ground measurements [@xavierDailyGriddedMeteorological2016]. After we fit vital rate models using the long-term survey data (@fig-lifecycle), we modeled delayed effects of SPEI with Distributed Lag Non-linear Models (i.e., DLNMs) with a maximum lag of 36 months [@scottDelayedEffectsClimate2022]. To iterate these parameter-resampled IPMs [*sensu* @metcalfStatisticalModellingAnnual2015] we first created a random sequence of SPEI values by sampling years of the observed (monthly) SPEI data. For every year we then calculated a lag of 36 months from the month in which that year's census was completed. These values were then used to predict the fitted values from vital rate models, which generated different kernels for each iteration of the IPM. The kernels of successive iterations are not entirely independent --- the SPEI values used to calculate vital rates include values used in the previous two iterations --- but they are ergodic.\

@@

We used a Generalized Linear Model to compare the projections of $\lambda$ from the two stochastic IPMs. We modelled the lambda as a function of IPM type (kernel-resampling vs parameter-resampling) and Habitat (continuous forest vs. fragments), and the interaction of IPM type and Habitat. Because the response variable ($\lambda$) was continuous we used the Gaussian distribution with the identity link function in our model. We verified the model assumptions by plotting residuals versus fitted values.\

@@

Finally, we compared the structure of populations through the first 250 time steps in each habitat by each of the stochastic IPMs (i.e, kernel- vs. parameter-resampling). At each time step, we assigned the individuals to one of four stage classes based on plant size, probability of survival, and probability of reproduction (see Table @tbl-plantcats). We then calculated the proportion of the population that was in each stage class in each habitat for each IPM type. Comparing the means and variances of these proportions will allow us to determine (1) if the two IPMs project different population structures over time, (2) if the population structure for a given IPM is consistent across habitat type. Because Shapiro-Wilk tests indicated that the distributions of proportions were not normally distributed, we used the non-parametric Ansari-Bradley Test to compare the distribution of proportions in each habitat x stage combination. These analyses were conducted using the R packages `rstatix` and `vartest`, respectively [@kassambaraRstatixPipefriendlyFramework2023; @cosarVartestTestsVariance2024].

@@

All IPMs projected higher population growth rates for *Heliconia acuminata* populations in Continuous Forest than for those in Forest Fragments (@tbl-lambdas, @tbl-glm). The differences between $\lambda^{CF}$ and $\lambda^{FF}$, which ranged from 1.19-2%, were significant for both deterministic ($P[\theta \geq \theta_{obs}]$ = `r det_stats$p_val`, N = 1000 randomization) and stochastic IPMs (@tbl-glm). There were also significant effects of IPM Type, Habitat, and their interaction on $\lambda$ (@tbl-glm). Because stochasticity is predicted to reduce population growth rates [@doakCorrectlyEstimatingHow2005a; @tuljapurkarManyGrowthRates2003; @metcalfStatisticalModellingAnnual2015], we were surprised to fine that deterministic $\lambda$ was nearly identical to the average $\lambda$ from kernel-resampled IPMs (@fig-lambdas). However, we did see reductions in population growth rate when including lagged effects in IPMs: on average the projections of $\lambda$ from these models were 1.5-2% lower, with 70-76% of the values below deterministic $\lambda$ (@fig-lambdas). The IPMs with lagged effects also appear to be more accurate --- the underlying vital rate models for IPMs with lagged effects all had the best fit to the survey data (dAIC = 0, @tbl-aic). <!--# No longer true if random effect = different smooth for each year -->

@@

What are the implications of these results for the use of IPMs to study plant demography? First, they suggest that failing to include lagged effects in models could significantly underestimate the growth rate of populations. While relative differences in projections of $\lambda$ may be sufficient for some tests of theory, the actual value of projections may be critical in conservation or management settings. For example, they could be used to identify which populations face the greatest risk of extinction (e.g., $\lambda=0.99$ vs. $\lambda=1.01$) or --- because even small differences in $\lambda$ can have major short-term consequences for population dynamics (e.g., $\lambda=1.01$ vs. $\lambda=1.02$) - be used to chose among management strategies. Moreover, including lagged effects in models could be especially critical for studies assessing the effects of climate change, given many DLHE are climate driven. That said, there is a cost to including lagged effects in demographic models, even when doing so is conceptually appropriate or if the outcome is potentially more accurate. While Deterministic and Kernel-resampled Stochastic IPMs took only \~`r det_time` and \~`r sker_time` min to iterate (respectively), the Stochastic IPMs with using parameter-resampled kernels and lagged effects took \~`r spar_time` min. This is largely due to the required computational resources and algorithms (i.e., `predict()`), which are much slower for Generalized Additive Models with 2D smooths because of the much higher number of knots than for the GAMs used in Determinsitic and Kernel-resampled IPMs. Advances in computational power and access to high-performance computing resources could lower this cost.

That said, computational power cannot compensate for limited data. Detecting lagged effects and evaluating their consequences requires long-term demographic data --- data that are only available for a relatively small number of species, few of which are in the tropics. The increasing evidence that lagged effects are ubiquitous, and that they can have major demographic impacts, underscores the need to support the collection of such long-term data, the complementary development of experimental and statistical approaches to disentangling lagged effects, and community driven efforts to identify priority or model systems for in-depth investigation.\

@@

<!-- Figure 2 Guides = collect not working?  -->

<!-- TODO: see notes table 5 -->

@@
